---
title: "Analysis of the Worth of Attending College in 2019"
output: html_notebook
---

##S.R. Mazeski <br /> 
###Senior Computational Data Science student at The Pennsylvania State University
stephenmazeski@gmail.com <br /> 
stephenmazeski.com



```{r}
library(mosaic)
library(tidyverse)
library(DataComputing)
library(lubridate)



```

```{r}
# Data has been hard imported into Avg_CPI
# CPI is the Consumer Price Index for the USA. Below is the head of the data along with an explination of its importance in this analysis.

head(Avg_CPI)

```


```{r}
# Log Graph of CPI highlights change in Average CPI.
ggplot(data=Avg_CPI,aes(x=YEAR,y=AVE.))+geom_point()+scale_y_log10() + geom_smooth() + labs(title = "US Avg CPI History", caption = "Based on data from inflationdata.com")

```
"The Consumer Price Index (CPI-U) is said to be based upon a 1982 Base for ease of thought. But in actuality the BLS set the index to an average for the period from 1982 through 1984 (inclusive) equal to 100, so there is no specific date when it was exactly 100. (If you check the chart below you will see it was somewhere between July and August of 1983.)
Note: Due to the width of the table, this page is best viewed full screen or as wide as possible." Via inflationdata.com

We can use this as a tool to guage the worth of USD over the course of the last century.
```{r}
head(WageGap)
```
```{r}
#with and without education


g <- ggplot(data = WageGap) + 
  geom_point(aes(x = Year, y = Lowest,colour = "Lowest")) +
  geom_point(aes(x = Year, y = Second, colour = "Second"))+
  geom_point(aes(x = Year, y = Third, colour = "Third")) +
  geom_point(aes(x = Year, y = Fourth, colour = "Fourth"))+
  geom_point(aes(x = Year, y = Fifth, colour = "Fifth"))

g + ylab("Household Income by Class") + scale_shape_discrete() + labs(title = "Wages by Five Distinct Classes in the US", caption = "Based on data from The US Census Bureau")+ scale_color_discrete(breaks=c("Fifth", "Fourth", "Third", "Second", "Lowest"))
```

```{r}
Income_Per_Ed <- Income_Per_education %>% select(Quentile, Less.than.High.School, High.School, Some.College.or.Associates, Bachelor.s.Degree..)
head(Income_Per_Ed)

```

```{r}
Income_Per_Ed$Quentile <- factor(Income_Per_Ed$Quentile, levels=c("Lowest", "Second", "Third", "Fourth", "Fifth"))
g2 <- ggplot(data = Income_Per_Ed) + 
  geom_point(aes(x = Quentile, y = Less.than.High.School,colour = "Less than HS")) +
  geom_point(aes(x = Quentile, y = High.School, colour = "High School"))+
  geom_point(aes(x = Quentile, y = Some.College.or.Associates, colour = "Some College/Associates")) +
  geom_point(aes(x = Quentile, y = Bachelor.s.Degree.., colour = "Bachelors Degree +"))

g2 + ylab("Number of People") + scale_shape_discrete() + labs(title = "Income Quentile by Education Level", caption = "Based on data from The bls.gov")+ scale_color_discrete(breaks=c("Less than HS", "High School", "Some College/Associates", "Bachelors Degree +"))


```

Clearly by this graphic, the cieling of earnings and number of jobs that allow people to climb to such wages are aided with further education.





















