---
title: "Analysis of the Worth of Attending College in 2019"
output: html_notebook
---

##S.R. Mazeski <br /> 
###Senior Computational Data Science student at The Pennsylvania State University
stephenmazeski@gmail.com <br /> 
stephenmazeski.com



```{r}
library(mosaic)
library(tidyverse)
library(DataComputing)
library(lubridate)
#A 52
#B 191
#F 50
#G 123
#H 105


#A 39, 60, 77
#B 39/60

```

```{r}
# Data has been hard imported into Avg_CPI
# CPI is the Consumer Price Index for the USA. Below is the head of the data along with an explination of its importance in this analysis.

head(Avg_CPI)

```


```{r}
# Log Graph of CPI highlights change in Average CPI.
ggplot(data=Avg_CPI,aes(x=YEAR,y=AVE.))+geom_point()+scale_y_log10() + geom_smooth() + labs(title = "US Avg CPI History", caption = "Based on data from inflationdata.com")

```
"The Consumer Price Index (CPI-U) is said to be based upon a 1982 Base for ease of thought. But in actuality the BLS set the index to an average for the period from 1982 through 1984 (inclusive) equal to 100, so there is no specific date when it was exactly 100. (If you check the chart below you will see it was somewhere between July and August of 1983.)
Note: Due to the width of the table, this page is best viewed full screen or as wide as possible." Via inflationdata.com

We can use this as a tool to guage the worth of USD over the course of the last century.

I think that the graph looks messy however with the arbituary AVE. and numbers from ~9 to 300.

I will use data wrangling in order to change this dataframe, and recreate the graphic.
```{r}
New_CPI <- Avg_CPI %>% mutate(CPI_Change = AVE./100)
head(New_CPI)


```
We now havea scalar value that may add more context to the extreme change in CPI.
Lets regraph the data.

```{r}
ggplot(data=New_CPI,aes(x=YEAR,y=CPI_Change))+geom_point()+scale_y_log10() + geom_smooth() + labs(title = "US CPI History", caption = "Based on data from inflationdata.com")

```

Lets see if we can create a graphic that highlights the difference in CPI by year.
```{r}
AveVec <- Avg_CPI$AVE.
DifferenceInCPI <- diff(AveVec)
revDiff <- rev(DifferenceInCPI)



```

```{r}
barplot(revDiff, main = "Change in CPI", xlab ="Yearly Change", col = ifelse(revDiff<0, 'red', 'green'))

```




```{r}
head(WageGap)
```
```{r}
#with and without education


g <- ggplot(data = WageGap) + 
  geom_point(aes(x = Year, y = Lowest,colour = "Lowest")) +
  geom_point(aes(x = Year, y = Second, colour = "Second"))+
  geom_point(aes(x = Year, y = Third, colour = "Third")) +
  geom_point(aes(x = Year, y = Fourth, colour = "Fourth"))+
  geom_point(aes(x = Year, y = Fifth, colour = "Fifth"))

g + ylab("Household Income (x100)") + scale_shape_discrete() + labs(title = "Wages by Five Distinct Classes in the US", caption = "Based on data from The US Census Bureau")+ scale_color_discrete(breaks=c("Fifth", "Fourth", "Third", "Second", "Lowest"))
```

```{r}
Income_Per_Ed <- Income_Per_education %>% select(Quentile, Less.than.High.School, High.School, Some.College.or.Associates, Bachelor.s.Degree..)
head(Income_Per_Ed)

ChangeInGap <- c(WageGap$Fifth - WageGap$Lowest)

```

```{r}
revChange <- rev(ChangeInGap)

```

```{r}
barplot(revChange, main = "Change in Wage Gap", xlab ="Yearly Change", col = ifelse(revDiff<0, 'red', 'green'))


```
This graphic highlights the change in the wage gap from the lowest 20% of Americans versus the top 20% of Americans in household incomes.


```{r}
Income_Per_Ed$Quentile <- factor(Income_Per_Ed$Quentile, levels=c("Lowest", "Second", "Third", "Fourth", "Fifth"))
g2 <- ggplot(data = Income_Per_Ed) + 
  geom_point(aes(x = Quentile, y = Less.than.High.School,colour = "Less than HS")) +
  geom_point(aes(x = Quentile, y = High.School, colour = "High School"))+
  geom_point(aes(x = Quentile, y = Some.College.or.Associates, colour = "Some College/Associates")) +
  geom_point(aes(x = Quentile, y = Bachelor.s.Degree.., colour = "Bachelors Degree +"))

g2 + ylab("Weekly Earnings") + scale_shape_discrete() + labs(title = "Earnings Described by Education Level", caption = "Based on data from The bls.gov")+ scale_color_discrete(breaks=c("Less than HS", "High School", "Some College/Associates", "Bachelors Degree +"))


```

Clearly by this graphic, the cieling of earnings and number of jobs that allow people to climb to such wages are aided with further education.



What if we can use an Auto-Regressive Moving Average to predict the CPI of the future?
```{r}
## First we implement a random walk ARIMA model
A_Data <- Avg_CPI$AVE.
B_Data <- Avg_CPI$YEAR
Future_CPI <- arima(A_Data, order = c(0,1,0))
m <- predict(Future_CPI, n.ahead = 10)
```

```{r}
m$se

```

This doesn't look very good to me. It seems as though the model predicts very inconsistently and when we refer back to the data, we can see this is true.

```{r}
head(Avg_CPI)

```
```{r}
## Now we implement a random walk ARIMA model with autocorrelation. Here we lag by 1 and autocorrect in order to create a more likely model.
A_Data <- Avg_CPI$AVE.
B_Data <- Avg_CPI$YEAR
Future_CPI <- arima(A_Data, order = c(1,1,0))
m2 <- predict(Future_CPI, n.ahead = 10)
```


```{r}
m2

```


```{r}
### Keep in mind we are at a CPI of 255.15 today.

TodayCPI = 255.15
PredictedCPI =c(1.5528,3.3147,5.3092,7.4504,9.6822,11.9657,14.2735,16.5855,18.8872,21.1679)
 #i = 0
 PredVector <- rep("", times=length((PredictedCPI)))
 for(i in seq_along(PredictedCPI)){
   holder <- (PredictedCPI[i]+TodayCPI)
   PredVector[i]<-holder
}
 PredVector
```

Here we now have predicted CPI values for the future based on ARIMA time-series analysis.


```{r}
JoinedDF <- merge(Income_Per_Ed, WageGap)

```
```{r}
head(JoinedDF)
```


```{r}
ggplot(data=JoinedDF,aes(x=Year, shape = Quentile))+geom_point(aes(y = High.School, color = 'High School'))+geom_point(aes(y = Less.than.High.School ,color= 'Less than HS'))+ geom_point(aes(y = Some.College.or.Associates, color = 'Some College/Associates'))+ geom_point(aes(y = Bachelor.s.Degree.., color = 'Bachelors Degree +')) + ylab("Weekly Income") + scale_color_discrete(breaks=c("Less than HS", "High School", "Some College/Associates", "Bachelors Degree +"))

```



